Active Embeddings: Conservatively Extending a Type System From Within

### 2014-03-01 att-icfp14/att-icfp14.tex
Exposition finally coming toghetr....

@@

\title{Active Typing: Conservatively Extending a Type System From Within}

@@

We introduce @$\lambda$, a simply-typed lambda calculus with simply-kinded type-level computation where new indexed type and operator constructors can be declared  from within. %It is structured in four layers: at the top-level, extension providers declare new indexed type constructors and term-level operator constructors. 
Type-level functions associated with operator constructors define their static and dynamic semantics, the latter by translation to a fixed typed internal language. By lifting compiler correctness techniques pioneered by the TIL compiler for Standard ML into the language, the ``actively typed semantics'' guarantees type safety. Going further, the semantics enforce a novel form of representation independence at extension boundaries that ensures that extensions are mutually conservative (i.e. they do not weaken or interfere with one another, so that they can  be used together in any combination). 
%Totality of the type-level language guarantees decidability of type checking and equality constraints on type indices makes type equivalence decidable. 
We intend @$\lambda$ as a minimal foundation for future work on safe language-integrated extension mechanisms for typed programming languages, but it is already quite expressive. We demonstrate by showing how a conventional concrete syntax can be introduced by type-directed dispatch to Core @$\lambda$, then discuss a number of typed language fragments that can be introduced as orthogonal ``libraries'' (and discuss the sorts of fragments that, as yet, cannot).
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}
Typed programming languages are often described in fragments, each consisting of a small number of indexed type constructors (often one) and associated (term-level) operator constructors. The simply typed lambda calculus (STLC), for example, consists of a single fragment containing a single type constructor, $\rightarrow$, indexed by a pair of types, and two operator constructors: $\lambda$, indexed by a type, and \verb|ap|, which we may think of as being indexed trivially. A fragment is often identified by the type constructor it is organized around, so the STLC might also be called $\mathcal{L}\{\rightarrow\}$, following the notational convention used by Harper \cite{pfpl} and others.

G\"odel's $\mathbf{T}$ consists of the $\rightarrow$ fragment and the \verb|nat| fragment, defining natural numbers and a recursor that allows one to ``fold'' over a natural number. We might thus call it $\mathcal{L}\{\rightarrow \mathtt{nat}\}$. This language is  more powerful than the STLC because the STLC admits an embedding (trivially) into  $\mathbf{T}$ but the reverse is not true. Buoyed by this fact, we might go on by adding fragments that define sums, products and various forms of inductive types (e.g. lists), or perhaps a general mechanism for defining inductive types. Each fragment clearly increases the expressiveness of our language.

If we consider the $\forall$ fragment, however, defining universal quantification over types (i.e. parametric polymorphism), we might take pause. In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girardF} and Reynolds as the polymorphic lambda calculus \cite{polylam}, it is known that sums, products, and inductive and co-inductive types can all be weakly defined via Church-style encodings \cite{reynolds}\todo{what to cite for this?}. This fact is of substantial theoretical importance and is quite relevant if \emph{translating} a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$ to $\mathcal{L}\{\rightarrow\forall\}$ is one's main interest. But Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, cautioned against using the existence of a weak encoding as an argument for programming with the corresponding embedding directly \cite{Reynolds94anintroduction}: 
\begin{quote}
To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
\end{quote}

@@

%Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language, 
Modern ``general-purpose'' languages like ML, on the other hand, expose constructs, like datatypes and abstract types, that admit strong embeddings of many simpler fragments, occupying what is widely seen as a ``sweet spot'' in the design space. However, ``general-purpose'' and ''all-purpose'' are quite distinct, and situations continue to arise in both research and practice where introducing a new fragment would be desirable because it can still only be weakly defined in terms of existing fragments, or a strong embedding, while possible, is impractically verbose or inefficient:
\begin{enumerate}
\item General-purpose abstractions continue to evolve and admit variants. There are many  variations on record types, for example: records with functional record update, specified field ordering (that is, labeled tuples) or prototypic delegation are either awkward to strongly embed or cannot be. Similarly, sum types also admit variants (e.g. various forms of open sums). Even something as seemingly simple as a type-safe \verb|sprintf| operator requires extending a language like Ocaml  \cite{ocaml-printf}.

\item Perhaps more interestingly, specialized type systems that enforce stronger invariants than general-purpose constructs are capable of enforcing are often developed by researchers. One need take only a brief  excursion through the literature to discover language extensions that support parallel programming \cite{a}\todo{citations}, concurrent programming \cite{cml}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow security \cite{walker00, smith2001}, database queries \cite{db}, aliased references \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and many others.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
\item Foreign function interfaces (FFIs) that allow programmers to safely and naturally interact with libraries written in an existing language require enforcing the type system of the foreign language within the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
Safe FFIs generally require direct extensions to the language. For example, MLj extended Standard ML with constructs for safely and naturally interfacing with Java \cite{mlj}. For any other language, including  others on the JVM like Scala and the many languages that might be accessible via Java's native FFI, there is no way to guarantee that language-specific invariants are statically maintained, and the interface is certainly far from natural.
\end{enumerate}

These sorts of innovations are generally disseminated as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, using tools like compiler generators, DSL frame\-works or  language workbenches, or directly within a particular compiler for the language, sometimes activated by a flag or pragma. This, we argue, is quite unsatisfying: a programmer can choose either a dialect supporting an innovative approach to parallel programming or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both. Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound language fragments remain sound when combined. Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so in a diverse fragment ecosystem, fragment providers have little choice but to leave this to clients. Avoiding language composition is difficult, as interoperability between components of an application written in different dialects faces precisely the problems described in item 3.% We will consider this further in our discussion of related work in Sec. \ref{related-work}.

@@

\paragraph{Contributions}
We draw inspiration from this approach to design a mechanism that makes it possible to declare new type and operator constructors and define, \emph{using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language. If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a form that can be implemented by this protocol, it can be made strongly embeddable in the external language by introducing, in this way, the necessary constructors and static logic.

The semantics imposes checks on extensions to guarantee type safety and decidability of typing and type equality. Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong embeddings of a language fragment can be derived assuming a ``closed world'' (that is, in the completely traditional manner where one constructs a complete language and reasons inductively). The semantics guarantees that they will be \emph{conserved} in the  ``open world'' by enforcing a form of  representation independence at all extension boundaries. This avoids the biggest problems of composing language dialects and justifies the inclusion of the mechanism directly into the language.

We will begin in Sec. \ref{core} by introducing a minimal calculus with this kind of \emph{actively typed semantics}, @$\lambda$. The external language begins with only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products. We implement these as extensions to explain the mechanism and motivate the constraints that the semantics imposes. We then examine the semantics in greater detail and state and sketch the key points in the proofs of the metatheory in Sec. \ref{theory}. These are the fundamental contributions of this work.

While the core calculus addresses the issue of making a strong embedding possible, embeddings into the core calculus are quite clearly impractical syntactically. To begin to address the issue of syntax, we develop in Sec. \ref{expanded-syntax} a flexible type-directed dispatch protocol for a more conventional \emph{expanded syntax} that defers to the core calculus semantically. We then use the expanded syntax to discuss several more interesting examples in Sec. \ref{examples}: \todo{list these}. \todo{organization of rest of sections}.

While of surprising expressiveness, @$\lambda$ itself is certainly not capable of expressing the full variety of type systems listed earlier. Indeed, to guarantee safety and conservativity, our calculus assumes perhaps too much about the type system being encoded. For example, it only admits fragments for which the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment; new contexts cannot be introduced. We discuss this and some of the other reasons why a fragment may not be expressible in @$\lambda$ in Sec. \ref{limitations}. Fortunately, these limitations do not appear fundamental to the approach, so we also outline in Sec. \ref{limitations} directions for future work. We conclude with a discussion of related work in Sec. \ref{related-work}.

@@

The syntax of Core @$\lambda$ is given in Fig. \ref{grammar}. An example of a program defining type and operator constructors that can be used to strongly embed G\"odel's \textbf{T} in @$\lambda$ is given in Fig. \ref{nat}. We will discuss its semantics and how precisely the strong embedding, seen being used starting on line 16, works as we go on. Natural numbers can, of course, be strongly embedded in existing languages with a similar usage and asymptotic performance profile (up to function call overhead with an abstract type, for example). We will provide more sophisticated examples where this is less feasible later on. %It consists of:
%a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}). Following the type constructor declaration, we apply $\fvar{Nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$. Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two. We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.

\subsection{Programs}
\begin{figure}[t]
\small
$$\begin{array}{rccl}	
\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort}
\\& &  \pipe & 
%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
 e\\
		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
												\topp{\theta}{\theta}\\
\\
\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
%														\efix{x}{\tau}{e} \pipe 
														\elam{\evar{x}}{\tau}{e} \pipe 
														\eop{Tycon}{op}{
															\tauidx
														}{
  												    		\splat{e}{1}{n}
														} \\
									& 		&		& 	\\

@@

\caption{\small Syntax of Core \atlam. Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over standard total binary operations over integers (e.g. addition).%The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.

@@

&}{XXX}{i}{\tlam{idx}{\kunit}{\titype{\dint}}};\\

@@

\caption{\small G\"odel's $\mathbf{T}$ in \atlam, used to calculate 2+2. The simple helper functions $\tvar{arity}n$ and $\tvar{ifeq}$ have return kind $\ksum{\kDen}{\kunit}$ (see text). Their definitions can be found in the appendix. We use \textsf{let} to bind both external and type-level variables for clarity of presentation; these can be eliminated manually. We will discuss the issue of a more natural syntax in Sec. \ref{expanded-syntax}.}

@@

A \emph{program}, $\rho$, consists of a series of static declarations (in the core, only constructor declarations) followed by an external term, $e$. The syntax for external terms contains three term formers: variables, $\lambda$, and a form for all other operator invocations, which we will discuss below. Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms (Fig. \ref{kindprog}, discussed below) then \emph{type checking} the external term and, simultaneously, \emph{translating} it to a term, $\iota$, in the {typed internal language} (Fig. \ref{att}). The key judgement is the \emph{active typing judgement}, relating an external term to a {type} and an internal term, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$: 
$$\ecompilesX{e}{\tau}{\iota}$$

This judgement follows standard conventions for the typing judgement of a simply-typed lambda calculus in most respects. The typing context $\Gamma$ maps variables to types and obeys standard structural properties. The key novelties here are 1) that the available type and operator constructors are not fixed by the language, but rather are tracked by the constructor context, $\fvalCtx$, and 2) that a translation is simultaneously derived. There is no separate operational semantics for the external language; the dynamic semantics are defined by this translation to the internal language (which might have a more conventional operational semantics). \todo{cite Harper-Stone / denotational semantics} The two contexts and the external term can be thought of as input to the judgement, while the type and translation are output (we do not consider a bidirectional approach \cite{LovasPfenning}\todo{cite} or type inference in this work, though this is likely a promising avenue for future research).

In our calculus, the internal language (IL), consisting of internal terms $\iota$ and internal types $\sigma$, provides partial functions (via the generic fixpoint operator of Plotkin's PCF \cite{PCF}), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines). In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined (e.g. Standard ML or CakeML \cite{cakeml}, leaving translation to a simpler IL to later stages of the compilation process).

\subsection{Types and Type Constructors}
\begin{figure}[t]
\small
$\fbox{\inferrule{}{\progOKX{\progsort}}}$
%~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$~~~
%$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
\begin{mathpar}
\inferrule[k-tycon]{
	\fvar{Tycon} \notin \text{dom}(\fvalCtx)\\
	\kEq{\kappaidx}\\
	\tKind{\tvarCtx}{\fvalCtx}{\taurep}{\karrow{\kappaidx}{\kIType}}\\
	\opType{\tvarCtx}{\fvalCtxX{\fvalDf}}{\theta}{\Theta}\\
	\progOK{\tvarCtx}{\fvalCtxX{\fvalDf}}{\rho}
}{
	\progOKX{\pfam{\familyDf}{\rho}}
}

@@

\inferrule[hide-trans]{
	\fvar{Tycon} \notin \Xi
}{
	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^{\checkmark}}}{\trepof{\ttype{Tycon}{\tauidx}}}
}
\end{mathpar}
\caption{\small Abstracted internal typing}
\label{ait}
\end{figure}
@$\lambda$ is an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation. Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus. The classifiers of terms at the type level are called \emph{kinds}, $\kappa$ to distinguish them from  \emph{types}, which are just one kind  of type-level value. We will write the kind of types as $\kTypeBlur$, though it is also written $\star$, \verb|T| or \verb|Type| in various similarly structured languages (see Sec. \ref{related-work}). The kinding rules for type-level terms are given in Fig. \ref{todo}\todo{refer}.% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions. In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.

>>>>

\newcommand{\compresslist}{
  \vspace{-1em}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}

@@

\title{Active Embeddings: Conservatively Extending a Type System From Within}

@@

We introduce @$\lambda$, a simply-typed lambda calculus with simply-kinded type-level computation where new indexed type and operator constructors can be declared  from within. %It is structured in four layers: at the top-level, extension providers declare new indexed type constructors and term-level operator constructors. 
Type-level functions associated with operator constructors define their static and dynamic semantics, the latter by translation to a fixed typed internal language. By lifting compiler correctness techniques into the language, the ``actively typed semantics'' guarantees type safety. Going further, the semantics enforce an abstraction barrier at extension boundaries that ensures that extensions are mutually conservative (i.e. they do not weaken or interfere with one another, so that they can  be used together in any combination). 
%Totality of the type-level language guarantees decidability of type checking and equality constraints on type indices makes type equivalence decidable. 
We intend @$\lambda$ as a minimal foundation for future work on safe language-integrated extension mechanisms for typed programming languages, but it is already quite expressive. We demonstrate by showing how a conventional concrete syntax can be introduced by type-directed dispatch to Core @$\lambda$, then discuss a number of typed language fragments that can be \emph{actively embedded} by this mechanism as orthogonal ``libraries''. %(and discuss the sorts of fragments that, as yet, cannot).
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}
Typed programming languages are often described in fragments, each consisting of a small number of indexed type constructors (often one) and associated (term-level) operator constructors. The simply typed lambda calculus (STLC), for example, consists of a single fragment containing a single type constructor, $\rightarrow$, indexed by a pair of types, and two operator constructors: $\lambda$, indexed by a type, and \verb|ap|, which we may think of as being indexed trivially. A fragment is often identified by the type constructor it is organized around, so the STLC is  called $\mathcal{L}\{\rightarrow\}$, following the notational convention used by Harper \cite{pfpl} and others.

G\"odel's $\mathbf{T}$ consists of the $\rightarrow$ fragment and the \verb|nat| fragment, defining natural numbers and a recursor that allows one to ``fold'' over a natural number. We might thus call it $\mathcal{L}\{\rightarrow \mathtt{nat}\}$. This language is  more powerful than the STLC because the STLC admits an embedding  into  $\mathbf{T}$ but the reverse is not true. Buoyed by this fact, we might go on by adding fragments that define sums, products and various forms of inductive types (e.g. lists), or perhaps a general mechanism for defining inductive types. Each fragment clearly increases the expressiveness of our language.

If we consider the $\forall$ fragment, however, defining universal quantification over types (i.e. parametric polymorphism), we must take a moment to reflect. In $\mathcal{L}\{\rightarrow\,\forall\}$, studied variously by Girard as System $\mathbf{F}$ \cite{girardF} and Reynolds as the polymorphic lambda calculus \cite{polylam}, it is known that sums, products, and inductive and co-inductive types can all be weakly encoded \cite{reynolds}\todo{what to cite for this?}. This means that we can \emph{translate} well-typed terms of a language like $\mathcal{L}\{\rightarrow\forall~\mathtt{nat} + \times\}$ to well-typed terms in $\mathcal{L}\{\rightarrow\forall\}$ in a manner that preserves their dynamic semantics. But Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, reminds us that programming with the corresponding \emph{embedding}, particularly one where only the dynamic semantics are preserved, may be unwise \cite{Reynolds94anintroduction}: 
\begin{quote}
To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
\end{quote}

@@

%Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language, 
\todo{revise per title}A strong embedding, i.e. an isomorphism that preserves static reasoning principles, in a sense that we will make more precise as we go on, is generally far more difficult to establish than a weak embedding. For example, the aforementioned embeddings into $\mathcal{L}\{\rightarrow\forall\}$ do not preserve type disequality and other equational principles. Harder still is establishing a strong embedding that also supports an acceptably natural programming style (measured, to first approximation, by the amount of boilerplate code needed) and a reasonable cost semantics.

Modern languages like ML do, of course, provide constructs, like datatypes and abstract types, that admit strong and often satisfying embeddings of many language fragments, occupying what is widely seen as a ``sweet spot'' in the design space. However, ``general-purpose'' and ''all-purpose'' remain quite distinct, and situations continue to arise in both research and practice where desirable fragments can still only be weakly defined in terms of general-purpose constructs, or a strong embedding, while possible, is widely seen as too verbose or inefficient. For example:

\begin{enumerate}
%\compresslist
\item General-purpose constructs continue to admit variants. There are many  variants of product types: records, records with functional record update, labeled tuples (records that specify a field ordering) and records with delegation are either awkward to strongly embed or cannot be. Similarly, sum types also admit many variants (e.g. various forms of open sums and object systems, seen in a certain light). Even something as seemingly simple as a type-safe \verb|sprintf| operator requires special support from a language, e.g. Ocaml \cite{ocaml-printf}.
\item Perhaps more interestingly, specialized type systems that enforce stronger invariants than general-purpose constructs are capable of enforcing are often developed by researchers. One need take only a brief  excursion through the literature to discover language extensions that support parallel programming \cite{a}\todo{citations}, concurrent programming \cite{cml}, distributed programming \cite{tom7}, dataflow programming \cite{reactiveml}, authenticated data structures \cite{popl13}, information flow security \cite{walker00, smith2001}, database queries \cite{db}, aliased references \cite{naden12}, network protocols \cite{sekar99}, units of measure \cite{keneddy} and many others.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
\item Safe and natural foreign function interfaces (FFIs) require enforcing the type system of the foreign language within the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
%Safe FFIs generally require direct extensions to the language. 
For example, MLj extended Standard ML with constructs for safely and naturally interfacing with Java \cite{mlj}. For any other language, including  others on the JVM like Scala and the many languages that might be accessible via a native FFI, there is no way to guarantee that language-specific invariants are statically maintained, and the interface is certainly far from natural.
\end{enumerate}

These sorts of innovations are generally disseminated as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, using tools like compiler generators, DSL frame\-works or  language workbenches, or directly within a particular compiler for the language, sometimes activated by a flag or pragma. This, we argue, is quite unsatisfying: a programmer can choose either a dialect supporting an innovative approach to parallel programming or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both. Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound dialects remain sound when combined. Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so in a diverse software ecosystem, fragment providers have little choice but to leave this task to clients (providing, perhaps, some informal guidelines or partial automation). Avoiding language composition is difficult, as interoperability between components of an application written in different dialects faces precisely the problems of item 3.% We will consider this further in our discussion of related work in Sec. \ref{related-work}.

@@

\paragraph{Contributions}
We draw inspiration from this approach to design a mechanism that makes it possible to declare new type and operator constructors and define, \emph{using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language. If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a form that can be implemented by this protocol, it can be made strongly embeddable in the external language by introducing, in this way, the necessary constructors and type assignment logic. We call such an embedding an \emph{active embedding}.

The semantics imposes checks on extensions to guarantee type safety and decidability of type assignment and equality. Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong embeddings of a language fragment can be derived in a suitable ``closed world'' (that is, in the traditional manner where one reasons inductively over a language composed of a minimal collection of fragments). The semantics guarantees that they will be \emph{conserved} in the  ``open world'' by enforcing a form of  representation independence at extension boundaries. This avoids the most fundamental problems of composing dialects and justifies the inclusion of the mechanism inside the language.

We will begin in Sec. \ref{core} by introducing a minimal calculus with this kind of {actively typed} semantics, @$\lambda$. The external language begins with only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even simple fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products. We implement these as extensions to explain the mechanism and motivate the constraints that the semantics imposes. We then examine the semantics in greater detail and state and sketch the key points in the proofs of the metatheory in Sec. \ref{theory}. These two sections are the fundamental contributions of this work.

While the core calculus addresses the issue of making a strong embedding possible, the embedding into the core calculus is quite clearly impractical syntactically. To begin to address the issue of syntax, we develop in Sec. \ref{expanded-syntax} a flexible type-directed dispatch protocol for a more conventional {expanded syntax} that defers to the core calculus semantically. We then use the expanded syntax to discuss several more interesting examples in Sec. \ref{examples}: labeled sums and products, the latter with delegation and functional field update, and a variant of \verb|sprintf|.\todo{update this if something doesn't make it}

While of surprising expressiveness given its minimality, @$\lambda$ itself is certainly not capable of admitting embeddings of the full variety of type system fragments enumerated earlier. Indeed, to guarantee safety and conservativity, our calculus assumes perhaps too much about the type system being encoded. For example, it only admits fragments where the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment, so new contexts cannot be introduced. We discuss this and some of the other reasons why a fragment may not be satisfyingly embeddable in Sec. \ref{limitations}. These limitations do not appear fundamental to the approach, so we also outline in Sec. \ref{limitations} directions for future work. We conclude with comparisons to related work in Sec. \ref{related-work}.

@@

The syntax of Core @$\lambda$ is given in Fig. \ref{grammar}. An example of a program defining type and operator constructors that can be used to strongly encode G\"odel's \textbf{T} in @$\lambda$ is given in Fig. \ref{nat}. We will discuss its semantics and how precisely the corresponding embedding, seen being used starting on line 16, works as we go on. Natural numbers can, of course, be strongly embedded in existing languages with a similar usage and asymptotic performance profile (up to function call overhead with an abstract type, for example). We will provide more sophisticated examples where this is less feasible later on (and discuss type abstraction itself in Sec. \ref{limitations}).\todo{if i don't have time/space, take this out} %It consists of:
%a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}). Following the type constructor declaration, we apply $\fvar{Nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$. Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two. We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.

\subsection{Overview}\label{programs}
\begin{figure}[t]
\small
$$\begin{array}{rccl}	
\textbf{programs} & \rho & ::= & \pfam{\familyDf}{\progsort}
\\& &  \pipe & 
%\pipe \pdef{t}{\kappa}{\tau}{\progsort} 
 e\\
		&	\theta	&	::= &	\tops{op}{\kappaidx}{i}{a}{\taut{def}} \pipe 
												\topp{\theta}{\theta}\\
\\
\textbf{external terms} 				&	e	&	::=	&	\evar{x} \pipe 
%														\efix{x}{\tau}{e} \pipe 
														\elam{\evar{x}}{\tau}{e} \pipe 
														\eop{Tycon}{op}{
															\tauidx
														}{
  												    		\splat{e}{1}{n}
														} \\
									& 		&		& 	\\

@@

\caption{\small Syntax of Core \atlam. Here, $x$ ranges over external and internal language variables, $\tvar{t}$ ranges over type-level variables, $\fvar{Tycon}$ ranges over type constructor names, $\opvar{op}$ ranges over operator constructor names, $\iintlit$ ranges over integer literals, $\tlabel{label}$ ranges over label literals (see text) and $\oplus$ ranges over standard total binary operations (e.g. addition, comparison).

@@

&}{XXX}{i}{\tlam{i}{\kunit}{\titype{\dint}}};\\

@@

\caption{\small An embedding of G\"odel's $\mathbf{T}$ in \atlam, used to define $plus$ and $two$ and calculate $plus~two~two$. The simple helper functions $\tvar{arity}n$ and $\tvar{ifeq}$ have return kind $\ksum{\kDen}{\kunit}$ and should be replaced by the definitions given in Fig. \ref{helpers}. We use \textsf{let} to bind both external and type-level variables for clarity of presentation. We will discuss further syntactic issues in Sec. \ref{expanded-syntax}.}

@@

A \emph{program}, $\rho$, consists of a series of static declarations (in the core, only constructor declarations) followed by an external term, $e$. The syntax for external terms contains three forms: variables, $\lambda$ terms, and a form for invoking user-defined operators, which we will discuss below. Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms, $\tau$, which we will describe in Sec \ref{types} (Fig. \ref{kindprog}), then typechecking the external term and simultaneously {translating} it to a term, $\iota$, in the {typed internal language}, described thereafter (Fig. \ref{att}). We can thus specify the \emph{central compilation judgement} $\pkcompiles{\rho}{\iota}$ with one rule:
\[
\inferrule[p-compiles]{
	\progOK{\emptyset}{\fvalCtx_0}{\rho}\\
	\pcompiles{\fvalCtx_0}{\rho}{\iota}
}{\pkcompiles{\rho}{\iota}}
\]
The key judgement form in the calculus is the \emph{active typing judgement}  (Fig. \ref{att}, which we will describe in Secs. \ref{opcons}-\ref{repind}), relating an external term to a {type}, $\tau$, called its \emph{type assignment}, and an internal term, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$: 
\[\ecompilesX{e}{\tau}{\iota}\]

The typing context $\Gamma$ is essentially conventional, mapping variables to types and obeying standard structural rules (weakening, exchange and contraction; \cite{pfpl} describes the necessary background for this paper). The constructor context tracks user-defined type and operator constructors, the topics of the remainder of this section, and establishing its structural properties requires establishing the conservativity properties that we will discuss in Sec. \ref{theory}.% Weakening and exchange of the constructor context closely related to the issue of conservativity that we will return to. Each constructor is identified by name, so there is no analog to contraction.

There is no separate operational semantics for the external language; the dynamic behavior of an external term is determined by its translation to the internal language (which might have a more conventional operational semantics). This form of semantics can be considered as a lifting into the language specification of the first stage of a type-directed compiler like the TIL compiler for Standard ML \cite{TIL} and has some parallels to the Harper-Stone semantics for Standard ML, where external terms were also given meaning by elaboration to an IL \cite{harper-stone}\todo{citations}. %

In @$\lambda$, the internal language (IL) provides partial functions (via the generic fixpoint operator of Plotkin's PCF), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines). In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined. Constraints that must be maintained for all well-typed terms, independent of the constructor context (e.g. guarantees that out-of-bounds access to memory never occurs), must be maintained by the internal type system and program performance is ultimately limited by the internal language and downstream compilation stages that we do not here consider (safely extending downstream compilation has been discussed in previous work, e.g. \cite{compexts}).

\subsection{Type Constructors and Types (Fig. \ref{nat}, lines 1 and 16)}\label{types}
\begin{figure}[t]
\small
$\fbox{\inferrule{}{\progOKX{\progsort}}}$
~~~$\fvalCtx ::= \emptyctx \pipe \fvalCtxX{\fvalDf}$
~~~$\tvarCtx ::= \emptyctx \pipe \tvarCtxX{t}{\kappa}$
%$\fCtx ::= \Sigma_0	 \pipe \fvalCtxX$
\begin{mathpar}
\inferrule[k-tycon]{
	\fvar{Tycon} \notin \text{dom}(\fvalCtx)\\
	\kEq{\kappaidx}\\
	\tKind{\tvarCtx}{\fvalCtx}{\taurep}{\karrow{\kappaidx}{\kIType}}\\
	\opType{\tvarCtx}{\fvalCtxX{\fvalDf}}{\theta}{\Theta}\\
	\progOK{\tvarCtx}{\fvalCtxX{\fvalDf}}{\rho}
}{
	\progOKX{\pfam{\familyDf}{\rho}}
}

@@

\inferrule[hide-trans]{
	\fvar{Tycon} \notin \Xi
}{
	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^{\checkmark}}}{\trepof{\ttype{Tycon}{\tauidx}}}
}
\end{mathpar}
\caption{\small Abstracted internal typing}
\label{ait}
\end{figure}
User-defined operator constructors are declared using \textsf{opcon} (cf. Fig. \ref{kindprog}).  For reasons that we will discuss, our calculus associates every operator  constructor with a type constructor. The \emph{fully-qualified name} of every operator constructor, e.g. $\fvar{Nat}.\opvar{z}$, must be unique. Operator constructors are indexed by type-level values, like type constructors, and so also declare an index kind, $\kappaidx$. In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but later examples will use more interesting indices. For example, the projection operator constructor for labeled tuples, $\fvar{LabeledTuple}.\opvar{prj}$, has index kind $\klabel$. An operator itself is, notionally, selected by indexing an operator constructor, e.g. $\fvar{Nat}.\opvar{s}\langle \tunit \rangle$, but note that neither operator constructors nor operators are first-class at any level (additional machinery would be needed, and this is not fundamental to our calculus).
